Binary Latent Representations for Efficient Ranking

### 2017-06-13 paper/paper.tex
Move related work to the end.
Small improvements.

To alleviate these constraints, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score and requiring a fraction of the memory. In MovieLens 100k experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take 55\% of the memory.

Our approach is complementary to approximate inner product search. Approximate MIPS proceeds in two stages: (1) identification of candidate result vectors through localicty-sensitive hashing, followed by (2) explicit scoring and sorting of the returned candidates. Replacing real-valued dot products by XNOR operations in the second stage will yield incremental performance gains over a real-valued MIPS approach.

\section{Our approach}
\label{sec:approach}
\citet{rastegari2016xnor} introduce XNOR-Networks, an approach that combines 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.

We adapt their approach to the recommendation task by learning binary versions of user and item vectors within a learning-to-rank factorization model. This allows us to use binary instead of floating point operations for computing recommendation rankings, which has significant speed and memory use advantages. In a real-valued $n$-dimensional factorization model, item and user representations take $4n$ bytes of memory, and computing a single prediction takes $2n$ floating point operations. In an equivalent binary model, $\frac{3}{32}n$ binary operations are required for each prediction (XOR, negation, and bitcounting), and the representations take $\frac{1}{8}n$ bytes of memory. Naive calculations suggest that, for the same latent dimensionality, a binary model would be over 20 times faster to score, and take less then 5\% of memory to store its representations.

@@

Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 100 repetitions of scoring 100,000 items and averaging the results.

\subsection{Implementation}
The model is implemented in PyTorch and trained using the nVidia K40 GPUs. During training, the embedding parameters are stored as floats. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule.

The prediction code runs on the CPU and is implemented using Intel X86 AVX2 SIMD intrinsics. SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version. We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of compiler autovectorization.

@@

The source code for the implementation is available on Github\footnote{\url{https://github.com/maciejkula/binge}}.

@@

Results on scoring speed and memory use broadly confirm the naive calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases.
As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 8\%, varying between 2\% and 16\%.
Nevertheless, the model's speed enables some very favourable tradeoffs.

When optimizing for MRR in both settings (at 1024 dimensions for the continuous and 2048 for the binary case), the XNOR representation achieves 96\% of the maximum MRR of the float representations at 13 times the query throughput and using only 6\% of the memory.

@@

>>>>

To alleviate these constraints, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score and requiring a fraction of the memory. In MovieLens 100k experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take 55\% of the memory. We make our model fitting and prediction code available on Github at \url{https://github.com/maciejkula/binge}.

\section{The model}
\label{sec:approach}
XNOR-Networks \citep{rastegari2016xnor} combine 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.

We adapt them to the recommendation task by learning binary versions of user and item vectors within a learning-to-rank factorization model. This allows us to use binary instead of floating point operations for computing recommendation rankings, which has significant speed and memory use advantages. In a real-valued $n$-dimensional factorization model, item and user representations take $4n$ bytes of memory, and computing a single prediction takes $2n$ floating point operations. In an equivalent binary model, $\frac{3}{32}n$ binary operations are required for each prediction (XOR, negation, and bitcounting), and the representations take $\frac{1}{8}n$ bytes of memory. Na{\"\i}ve calculations suggest that, for the same latent dimensionality, a binary model would be over 20 times faster to score, and take less then 5\% of memory to store its representations.

@@

Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 100 repetitions of scoring 100,000 items and averaging the results. Memory use is given by the total number of bytes required to store all the parameters of each model (including biases) and scaling parameters in the binary model.

\subsection{Implementation}
The model is implemented in PyTorch and trained using the nVidia K40 GPUs. During training, the embedding parameters are stored as floats. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule \citep{kingma2014adam}.

The prediction code runs on the CPU and is implemented using Intel X86 AVX2 SIMD intrinsics. SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version. We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of automatically-generated SIMD operations.

@@

@@

Results on scoring speed and memory use broadly confirm the na{\"\i}ve calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases.
As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 8\%, varying between 2\% and 16\%.

Nevertheless, the model's speed enables some favourable tradeoffs. When optimizing for MRR in both settings (at 1024 dimensions for the continuous and 2048 for the binary case), the XNOR representation achieves 96\% of the maximum MRR of the float representations at 13 times the query throughput and using only 6\% of the memory.

@@

Our method is complementary to approximate inner product search. Approximate MIPS proceeds in two stages: (1) identification of candidate result vectors through localicty-sensitive hashing, followed by (2) explicit scoring and sorting of the returned candidates. Replacing real-valued dot products by XNOR operations in the second stage will yield incremental performance gains over a real-valued MIPS approach.
